prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-06-07 inst/MyArticle/paper.Rmd
more suggestions from PN.

---
title: prolfqua: A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [Swiss Federal Institute of Technology in Zurich](https://ethz.ch) \| [University of Zurich](https://uzh.ch); Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB)](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
date: "`r format(Sys.time())`"
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

R linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. 
Models in `r BiocStyle::Githubpkg("fgcz/prolfqua")` are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending; @MASS2002] is advantageous when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` using base R or other programming languages by reading the analysis script. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of `r BiocStyle::Biocpkg("MSstats")`, we provide the functionality to suggest a model formula from the sample annotation provided in tabular form similarly to `r BiocStyle::Biocpkg("MSstats")` [@Choi2014].

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves and computes partial areas under those curves (pAUC) and other scores, e.g., the false discovery proportion FDP. We use those scores, i.e. the $pAUC$ at $10%$ FDR and the FDP, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with results produced by `r BiocStyle::Biocpkg("MSstats")` [@MSstats2014] and `r BiocStyle::Biocpkg("proDA")` [@bioxrvproDA2020]. Table \@ref(tab:describeBenchmarkedModels) summarizes all modelling methods we evaluated.

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimates differences (see Figure \@ref(fig:FDRfdp) Panel A). The number of proteins indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4046 proteins with at least two peptides, there are in total $16184$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as pAUC. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ group differences while the mixed effect models estimates the fewest with $15940$. Hence, to conclude that one method shows a better performance, it does not suffice that pAUC is greater, but the number of proteins with differential expression results needs to be equal or larger.

@@

There are only minor differences in the $pAUC_{10}$ between the $t$-statistics or the scaled $p$-value  (see Figure \@ref(fig:FDRfdp) Panel A). Interestingly, the $pAUC$ increases when using $p$-values instead of the $t$-statistics for linear models, while it decreases for mixed effect models. The reason is an erroneous denominator degrees of freedom estimation for many proteins in the case of the mixed-effect models.

@@

xMM <- data.frame(t(xMM))
rownames(xMM) <- NULL
colnames(xMM) <-  c("Label","Description","Abundance","Input File")
kableExtra::kable(xMM,
             caption = paste0("All benchmarked models.",
             "Description - prolfqua function names",
             "Abudances - indicates if model is fitted to peptide or protein abundances",
             "Input File - MaxQuant file used as input to the analysis."),
             format = "latex", booktabs=TRUE) |>
    kableExtra::kable_styling(latex_options="scale_down") |> 
    kableExtra::add_footnote(label = "Merges results of the prot_med_lm_moderated` and prot_imputation modeling pipeline, preferring those of prot_med_lm_moderated if available.")

@@

Since only technical replicates are available for each dilution, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. Furthermore, the dataset captures only the variance from the chromatography, electro-spray, and mass spectrometric measurement method. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret the results. Specifically, the observed variances will be higher in more realistic experiments, and the power will be lower for the same sample sizes. Furthermore, the proportion of missing observations in real-life datasets might also be higher.

We can conclude that if we want to order the proteins according to the likelihood of being differentially regulated to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might improve the inference. However, this improvement is minuscule (see Figure \@ref(fig:FDRfdp) panel B).
There is no such improvement for the mixed effect model, most likely because the degrees of freedom are erroneously estimated for many models. Furthermore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by [@Smyth2004linear], consistently improves the ranking of proteins compared with the unmoderated estimates (not shown) and fails to do so for mixed effect models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest $AUC$ scores among all the tested models (see Figure \@ref(fig:FDRfdp) Panel A left). Furthermore, by using the beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:FDRfdp) Panel A center). However, the properties of beta-based probabilities are not well understood; for instance, the p-values are not uniformly distributed under the null hypothesis (not shown). Furthermore, the FDR estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased and overestimate the false discovery proportion (see Figure \@ref(fig:FDRfdp) Panel C). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:FDRfdp) Panel C).

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A). Because of the robustness of the dropout model to missing observations we obtain difference estimates for all proteins and contrasts (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated differences are unbiased and show high diagnostic accuracy (\@ref(fig:FDRfdp) Panel A). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the linear model with variance moderation (`prot_med_lm_moderated`). Therefore, we started to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

The R-package `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Biocpkg("prolfqua")` model the missing data directly, while MSstats imputes the data using an accelerated failure model. Despite imputation, `r BiocStyle::Biocpkg("MSstats")` do not estimate group differences for more proteins and do not achieve a higher $pAUC$ score than `r BiocStyle::Githubpkg("fgcz/prolfqua")`. Furthermore, \@ref(fig:FDRfdp) Panel C shows that when using `r BiocStyle::Biocpkg("MSstats")` the proportion of false discoveries might be very high even when filtering for a low FDR.

>>>>

---
title: prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [Swiss Federal Institute of Technology in Zurich](https://ethz.ch) \| [University of Zurich](https://uzh.ch); Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB)](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
date: "`r format(Sys.time())`"
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

R linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. 
Models in `r BiocStyle::Githubpkg("fgcz/prolfqua")` are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending; @MASS2002] is advantageous when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` using base R or other programming languages by reading the analysis script. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of `r BiocStyle::Biocpkg("MSstats")`, we provide the functionality to suggest a model formula from the sample annotation provided in tabular form similarly to `r BiocStyle::Biocpkg("MSstats")`.

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves (ROC) and computes partial areas under those curves (pAUC) and other scores, e.g., the false discovery proportion FDP. We use those scores, i.e. the $pAUC$ at $10%$ FDR and the FDP, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with results produced by `r BiocStyle::Biocpkg("MSstats")` and `r BiocStyle::Biocpkg("proDA")`. Table \@ref(tab:describeBenchmarkedModels) summarizes all modelling methods we evaluated.

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimated differences (see Figure \@ref(fig:FDRfdp) Panel A). It indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4046 proteins with at least two peptides, there are in total $16184$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as pAUC. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ group differences while the mixed effect models estimates the fewest with $15940$. Hence, to conclude that one method shows a better performance, it does not suffice if the pAUC is greater, but the number of proteins with differential expression results needs to be equal or larger.

@@

There are only minor differences in the $pAUC_{10}$ between the $t$-statistics or the scaled $p$-value  (see Figure \@ref(fig:FDRfdp) Panel A). Interestingly, the $pAUC$ increases when using $p$-values instead of the $t$-statistics for linear models, while it decreases for mixed effect models. The reason is erroneous denominator degrees of freedom estimation for many proteins in the case of the mixed effect models.

@@

xMM <- data.frame(t(xMM))
rownames(xMM) <- NULL
colnames(xMM) <-  c("Label","Description","Abundance","Input File")
kableExtra::kable(xMM,
             caption = paste0("All benchmarked models. ",
             "Description - prolfqua function names, ",
             "Abudances - indicates if model is fitted to peptide or protein abundances, ",
             "Input File - name of MaxQuant file used as input."),
             format = "latex", booktabs=TRUE) |>
    kableExtra::kable_styling(latex_options="scale_down") |> 
    kableExtra::add_footnote(label = "Merges results of the 'prot_med_lm_moderated' and 'prot_imputation' modeling pipeline, preferring those of 'prot_med_lm_moderated' if available.")

@@

Since only technical replicates are available for each dilution, essential sources of variation present in every experiment, such as the biochemical and biological, are not measured. Therefore, the dataset captures only the variance from the chromatography, electro-spray, and mass spectrometric measurement method. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret the results. Specifically, the observed variances will be higher in more realistic experiments, and the power will be lower for the same sample sizes. Furthermore, the proportion of missing observations in real-life datasets might also be higher.

We can conclude that if we want to sort the proteins according to the likelihood of being differentially regulated to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might improve the inference. However, this improvement is minuscule (see Figure \@ref(fig:FDRfdp) panel B).
There is no such improvement for the mixed effect model, most likely because the degrees of freedom are erroneously estimated for many models. Furthermore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by @Smyth2004linear, consistently improves the ranking of proteins compared with the unmoderated estimates (not shown) and fails to do so for mixed effect models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest $AUC$ scores among all the tested models (see Figure \@ref(fig:FDRfdp) Panel A left). Furthermore, by using the Beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:FDRfdp) Panel A center). However, the properties of Beta-based probabilities are not well understood; for instance, the p-values are not uniformly distributed under the null hypothesis (not shown). Furthermore, the FDR estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased and overestimate the false discovery proportion (see Figure \@ref(fig:FDRfdp) Panel C). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:FDRfdp) Panel C).

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A). Because of the robustness of the dropout model to missing observations we obtain difference estimates for all proteins and contrasts (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated differences are unbiased and show high diagnostic accuracy (Figure \@ref(fig:FDRfdp) Panel A). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the linear model with variance moderation (`prot_med_lm_moderated`). Therefore, we are planning to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

The R-package `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Biocpkg("prolfqua")` model the missing data directly, while MSstats imputes the data using an accelerated failure model. Despite imputation, `r BiocStyle::Biocpkg("MSstats")` does not estimate group differences for more proteins and does not achieve a higher $pAUC$ score than `r BiocStyle::Githubpkg("fgcz/prolfqua")`. Furthermore, Figure \@ref(fig:FDRfdp) Panel C shows that when using `r BiocStyle::Biocpkg("MSstats")` the proportion of false discoveries might be very high even when filtering for a low FDR.
